Accelerating Optimal Power Flow with GPUs: SIMD Abstraction of Nonlinear Programs and Condensed-Space Interior-Point Methods

### 2024-02-19 paper/main.tex
highlight added

The adoption of graphics processing units (GPUs) in the mathematical programming community has remained
limited. Notably, nonlinear programming (NLP) remains dependent on
algorithms developed in the 1990s offering limited room for parallelism.
One of the primary challenges arises from the
automatic differentiation (AD) of sparse model equations and the
parallel factorization of indefinite sparse matrices, which are
commonly encountered within constrained numerical optimization tasks
\cite{anitescu2021targeting}. While GPU computation can trivially
accelerate several parts of the optimization process---especially
various internal computations within the optimization solver---the
sluggish data transfer between host and device memory hampers the
ad hoc implementation of GPU accelerations. To
fully leverage the potential offered by modern GPU hardware, a comprehensive computational framework for
optimization on GPUs is imperative. That is, we need an AD/algebraic
modeling framework, sparse linear solvers, and NLP solvers that can
operate entirely on the GPU. Specifically, for the best performance,
both the problem data and the solver's intermediate computational data
must be exclusively resident within the device memory, with the
majority of operations executed on the GPU.

@@

This paper presents our approach to implementing a comprehensive
computational framework for solving large-scale alternating current optimal power flow (ACOPF) problems on
NVIDIA GPUs, along with the associated software implementations:
ExaModels.jl \cite{simdiff}, an algebraic modeling/AD tool, and MadNLP.jl \cite{madnlp}, an
NLP solver. Our approach incorporates two novel strategies: (i) a
single-instruction, multiple-data (SIMD) abstraction of nonlinear
programs, enabling streamlined parallel AD on GPUs, and (ii) a
condensed-space interior-point method (IPM) with an inequality
relaxation strategy, which facilitates the use of highly efficient
{\it refactorization} routines for sparse matrix factorization with
fixed pivot sequences.


While derivative evaluation can be generally cheaper than linear
algebra operations, our numerical results on ACOPF problems show that
AD often constitutes more than half of the total solver time when
using off-the-shelf AD implementations such as
JuMP.jl~\cite{dunning2017jump} or
AMPL~\cite{fourer1990modeling}. Instead, our method leverages a
specialized AD implementation based on the SIMD abstraction of
NLPs. This abstraction allows us to preserve the parallelizable
structure within the model equations, facilitating efficient
derivative evaluations on the GPU. The AC power flow model is
particularly well suited for this abstraction because it involves
repetitive expressions for each component type (e.g., buses, lines,
generators) and the the number of computational patterns does not
increase with the network's size. {\color{blue} In other words, the
objective and constraints can be expressed in the form of iterators.}
Numerical results reported in this paper (Table \ref{tbl:results})
demonstrate that our proposed {\color{blue} AD strategy} can achieve
over 20 times speedup by running on the GPU. Compared with general AD
implementations on CPUs (such as AMPL and JuMP.jl), our GPU-based
differentiation method can be approximately 500 times faster.

@@

We present numerical benchmark results to showcase the efficiency of
our method, utilizing our two packages MadNLP.jl and
ExaModels.jl. The  KKT system is solved using the
external cuSOLVER library. To assess the performance of our method, we
compare it with standard CPU approaches using the data available in
pglib-opf \cite{babaeinejadsarookolaee2019power}.  Our benchmark results
demonstrate that our proposed computational framework has significant
potential for accelerating the solution of ACOPF problems, especially
when a moderate accuracy (e.g., $10^{-4}$) is sufficient.
Notably, when running on NVIDIA GPUs, our method achieves a
4x speedup compared with our solver running on CPUs for
the largest instance. Moreover, for the same instance, our approach
surpasses the performance of existing tools (such as Ipopt interfaced
with JuMP.jl) by an order of magnitude. This finding underscores the
importance of harnessing the power of GPUs for tackling
the computational challenges in power systems.

@@

\paragraph*{Related Work}
Several recent works have explored the use of GPUs for large-scale
nonlinear optimization problems. Cao et al. \cite{cao2016augmented}
proposed an augmented Lagrangian interior-point approach that employs
augmented Lagrangian outer iteration and the treatment of linear
systems using a preconditioned conjugate gradient method. Prior to the
introduction of the sparse condensed-space IPM with an inequality relaxation
strategy, the authors  investigated the use of reduction
strategies (state variable elimination) to treat KKT systems in a
dense form on the GPU
\cite{pacaud2023parallel,pacaud2022feasible,pacaud2023accelerating}.
In parallel, approaches based on Lagrangian decomposition and
batched with batched TRON solver \cite{lin1999newton}
have been investigated \cite{kim2022accelerated,kim2021leveraging}.
An NLP solver for high-performance computers  with GPU
accelerators, called HiOP, has been under development
\cite{hiop_techrep}, with a  scope similar to that of our solver MadNLP.jl.
 Another recent development is the hybrid (direct-iterative) KKT
system solver specifically designed for GPUs \cite{regev2023hykkt}.
The implementation of derivative evaluations with the exploitation of
repeated structures within model equations was, to the best of our
knowledge,  first introduced in Gravity \cite{Gravity}. This was
achieved through the introduction of so-called template constraints,
and multithreaded derivative evaluation has been implemented therein.
We note, however,  that their differentiation approach
is based on symbolic differentiation.
The idea of condensed-space IPM (without inequality relaxation
strategy) is not new, but it has been used primarily in more
specific contexts, where the increased nonzero entries in the KKT system
are less of a concern, such as in the dense form model predictive control
problems \cite{jerez2012sparse,cole2023exploiting}.

@@

(3) The optimization solver employs a filter line search procedure to
determine the step size~\cite{wachter2006implementation}. The primal-dual iterate is updated by
applying the determined step size and direction. This process is
repeated until  satisfaction of the convergence criteria (typically based on the residual
to the first-order optimality conditions).

@@

Implementing kernel functions for the above computational patterns in
the Julia language is straightforward, since Julia provides excellent
high-level interfaces for array and kernel programming for GPU
arrays. The code can even be device-agnostic, thanks to the portable
programming capabilities brought by KernelAbstractions.jl. All
of the AD and optimization capabilities in our
tools MadNLP.jl and ExaModels.jl are implemented in  Julia,
by leveraging its kernel and array programming capabilities.

@@

\subsection{Sparsity Analysis}
The sparsity analysis is needed for determining the sparsity pattern
of the evaluated derivatives. In the case of large-scale sparse
problems, the initial sparsity analysis of nonlinear expressions can
be expensive, since the sparsity should be analyzed for potentially
millions of objective and constraints terms. Often, however, 
these analyses are applied to the same computational patterns, and
the time and memory spent on sparsity analysis can be significantly
reduced if the repetitive structures are exploited.

@@

\ref{config-1} is our main GPU configuration; \ref{config-2}
represents our implementation running on CPU; and \ref{config-3} and
\ref{config-4} are used as benchmarks. \ref{config-1} and
\ref{config-2} share significant amount of code, especially the
high-level abstractions, but they differ in how they handle the KKT
systems. In \ref{config-1}, MadNLP.jl applies the condensed-space IPM
along with the inequality relaxation strategy, while in
\ref{config-2}, MadNLP.jl applies IPM based on the indefinite,
noncondensed KKT system, as in \eqref{eqn:kkt-indefinite}. In
\ref{config-1}, we use the cuSOLVER library to solve the condensed KKT
system via Cholesky factorization. The numerical
factorization and triangular solves are performed by cuSOLVER with the
fixed pivot sequence obtained with an approximate minimum degree ordering algorithm \cite{amestoyApproximateMinimumDegree1996}, implemented in AMD.jl \cite{montoisonAMDJlJulia2020}.  Software and hardware details of each
configuration are illustrated in Table \ref{tbl:settings}. The ACOPF
problem is formulated by using the model from the rosetta-opf project
\cite{rosetta-opf}, which is based on the models in PowerModels.jl \cite{8442948}, and the test cases are obtained from the pglib-opf
repository \cite{babaeinejadsarookolaee2019power}. We have selected
the goc and pegase cases because they contain large-scale instances.  The
external packages are called from Julia, through thin wrapper
packages, such as Ipopt.jl and AmplNLWriters.jl. A tolerance of
$10^{-4}$ is set for MadNLP.jl and Ipopt solvers, with other solver
options adjusted to ensure a fair comparison across different
solvers. The results can be reproduced with the script available at
\url{https://github.com/sshin23/opf-on-gpu}.

@@

\paragraph*{Convergence Pattern}
When comparing the solvers' performance in terms of the
IPM iteration counts, MadNLP.jl is as
efficient as the state-of-the-art solver Ipopt. The IPM iteration
count is nearly the same as that of Ipopt for achieving the same level
of accuracy in the final solution (see Table \ref{tbl:quality}).
{\color{blue} Note that
the constraint violation is not strictly less than the tolerance in all configurations
as relative constraint violation is used as convergence criteria.}
This
suggests that running mathematically equivalent operations on GPUs (by
using a shared code base in high-level abstractions) can yield a similar
degree of effectiveness in terms of IPM convergence.

@@

\paragraph*{Performance of the Condensed-Space IPM}
We next discuss the linear solver time. While the speedup achieved by
linear solvers is only moderate, this has a high impact on the overall
speedup because the linear solver time constitutes a significant
portion of the total solution time. {\color{blue} This can be observed from Figure
\ref{fig:speedup}; linear algebra is the only bottleneck in
large-scale instances.  For large instances, GPUs can only achieve
approximately 5x speedup.  These speedup ratios can vary depending on
the instances.  For instance, for case24464\_goc, the GPU linear
solver performance was close to that of the CPU solver.  The
investigation of under which circumstances cuSOLVER is more effective
warrants further research.}

@@

\paragraph*{Alternative Linear Solvers}
While cuSOLVER has been effective for solving the condensed KKT
systems using LU factorization, Cholesky factorization holds promise
for better performance due to lower computational complexity and the
ability to reveal the inertia of the KKT system. We are
interested in exploring other linear solvers options, such as CHOLMOD
\cite{chen2008algorithm}, Baspacho \cite{pineda2022theseus}, and HyKKT
\cite{regev2023hykkt}.

>>>>

The adoption of graphics processing units (GPUs) in the mathematical programming community has remained
limited due to the challenges associated with parallelizing the optimization algorithms. Notably, nonlinear programming (NLP) remains dependent on
algorithms developed in the 1990s offering limited room for parallelism.
One of the primary challenges arises from the
automatic differentiation (AD) of sparse model equations and the
parallel factorization of indefinite sparse matrices, which are
commonly encountered within constrained numerical optimization tasks
\cite{anitescu2021targeting}. While GPU computation can trivially
accelerate several parts of the optimization process---especially
various internal computations within the optimization solver---the
sluggish data transfer between host and device memory hampers the
ad hoc implementation of GPU accelerations. To
fully leverage the potential offered by modern GPU hardware, a comprehensive computational framework for
optimization on GPUs is imperative. That is, we need an AD/algebraic
modeling framework, sparse linear solvers, and NLP solvers that can
operate entirely on the GPU. Specifically, for the best performance,
both the problem data and the solver's intermediate computational data
must be exclusively resident within the device memory, with the
majority of operations executed on the GPU.

@@

This paper presents our approach to implementing a comprehensive
computational framework for solving large-scale alternating current optimal power flow (ACOPF) problems on
NVIDIA GPUs, along with the associated software implementations:
ExaModels.jl \cite{simdiff}, an algebraic modeling/AD tool, and MadNLP.jl \cite{madnlp}, an
NLP solver. Our approach incorporates two novel strategies: (i) a
single-instruction, multiple-data (SIMD) abstraction of nonlinear
programs, enabling streamlined parallel AD on GPUs, and (ii) a
condensed-space interior-point method (IPM) with an inequality
relaxation strategy, which facilitates the use of highly efficient
{\it refactorization} routines for sparse matrix factorization with
fixed pivot sequences.

While derivative evaluation can be generally cheaper than linear
algebra operations, our numerical results on ACOPF problems show that
AD often constitutes more than half of the total solver time when
using off-the-shelf AD implementations such as
JuMP.jl~\cite{dunning2017jump} or
AMPL~\cite{fourer1990modeling}. Instead, our method leverages a
specialized AD implementation based on the SIMD abstraction of
NLPs. This abstraction allows us to preserve the parallelizable
structure within the model equations, facilitating efficient
derivative evaluations on the GPU. The AC power flow model is
particularly well suited for this abstraction because it involves
repetitive expressions for each component type (e.g., buses, lines,
generators) and the the number of computational patterns does not
increase with the network's size. {\color{blue} In other words, the
objective and constraints can be expressed in the form of iterators.}
Numerical results reported in this paper (Table \ref{tbl:results})
demonstrate that our proposed {\color{blue} AD strategy} can achieve
over {\color{blue} 10} times speedup by running {\color{blue} the AD} on the GPU. Compared with general AD
implementations on CPUs (such as AMPL and JuMP.jl), our GPU-based
differentiation method can be approximately 500 times faster.

@@

We present numerical benchmark results to showcase the efficiency of
our method, utilizing our two packages MadNLP.jl and
ExaModels.jl. The  KKT system is solved using the
external cuSOLVER library. To assess the performance of our method, we
compare it with standard CPU approaches using the data available in
pglib-opf \cite{babaeinejadsarookolaee2019power}.  Our benchmark results
demonstrate that our proposed computational framework has significant
potential for accelerating the solution of ACOPF problems, especially
when a moderate accuracy (e.g., $10^{-4}$) is sufficient.
Notably, when running on NVIDIA GPUs, our method achieves a
4x speedup compared with our solver running on CPUs for
the largest instance. Moreover, for the same instance, our approach
surpasses the performance of existing tools (such as Ipopt interfaced
with JuMP.jl) by an order of magnitude. This finding underscores the
importance of harnessing the power of GPUs to tackle
the computational challenges in power systems.

@@

\paragraph*{Related Work}
Several recent works have explored the use of GPUs for large-scale
nonlinear optimization problems. Cao et al. \cite{cao2016augmented}
proposed an augmented Lagrangian interior-point approach that employs
augmented Lagrangian outer iteration and the treatment of linear
systems using a preconditioned conjugate gradient method. Before the
introduction of the sparse condensed-space IPM with an inequality relaxation
strategy, the authors  investigated the use of reduction
strategies (state variable elimination) to treat KKT systems in a
dense form on the GPU
\cite{pacaud2023parallel,pacaud2022feasible,pacaud2023accelerating}.
In parallel, approaches based on Lagrangian decomposition and
batched with batched TRON solver \cite{lin1999newton}
have been investigated \cite{kim2022accelerated,kim2021leveraging}.
An NLP solver for high-performance computers  with GPU
accelerators, called HiOP, has been under development
\cite{hiop_techrep}, with a scope similar to that of our solver MadNLP.jl.
 Another recent development is the hybrid (direct-iterative) KKT
system solver specifically designed for GPUs \cite{regev2023hykkt}.
The implementation of derivative evaluations with the exploitation of
repeated structures within model equations was, to the best of our
knowledge,  first introduced in Gravity \cite{Gravity}. This was
achieved through the introduction of so-called template constraints,
and multithreaded derivative evaluation has been implemented therein.
We note, however,  that their differentiation approach
is based on symbolic differentiation.
The idea of condensed-space IPM (without inequality relaxation
strategy) is not new, but it has been used primarily in more
specific contexts, where the increased nonzero entries in the KKT system
are less of a concern, such as in the dense form model predictive control
problems \cite{jerez2012sparse,cole2023exploiting}.

@@

(3) The optimization solver employs a filter line search procedure to
determine the step size~\cite{wachter2006implementation}. The primal-dual iterate is updated by
applying the determined step size and direction. This process is
repeated until the satisfaction of the convergence criteria (typically based on the residual
to the first-order optimality conditions).

@@

Implementing kernel functions for the above computational patterns in
the Julia language is straightforward since Julia provides excellent
high-level interfaces for array and kernel programming for GPU
arrays. The code can even be device-agnostic, thanks to the portable
programming capabilities brought by KernelAbstractions.jl. All
of the AD and optimization capabilities in our
tools MadNLP.jl and ExaModels.jl are implemented in  Julia,
by leveraging its kernel and array programming capabilities.

@@

\subsection{Sparsity Analysis}
The sparsity analysis is needed to determine the sparsity pattern
of the evaluated derivatives. In the case of large-scale sparse
problems, the initial sparsity analysis of nonlinear expressions can
be expensive, since the sparsity should be analyzed for potentially
millions of objective and constraint terms. Often, however, 
these analyses are applied to the same computational patterns, and
the time and memory spent on sparsity analysis can be significantly
reduced if the repetitive structures are exploited.

@@

\ref{config-1} is our main GPU configuration; \ref{config-2}
represents our implementation running on CPU; and \ref{config-3} and
\ref{config-4} are used as benchmarks. \ref{config-1} and
\ref{config-2} share a significant amount of code, especially the
high-level abstractions, but they differ in how they handle the KKT
systems. In \ref{config-1}, MadNLP.jl applies the condensed-space IPM
along with the inequality relaxation strategy, while in
\ref{config-2}, MadNLP.jl applies IPM based on the indefinite,
noncondensed KKT system, as in \eqref{eqn:kkt-indefinite}. In
\ref{config-1}, we use the cuSOLVER library to solve the condensed KKT
system via Cholesky factorization. The numerical
factorization and triangular solves are performed by cuSOLVER with the
fixed pivot sequence obtained with an approximate minimum degree ordering algorithm \cite{amestoyApproximateMinimumDegree1996}, implemented in AMD.jl \cite{montoisonAMDJlJulia2020}.  Software and hardware details of each
configuration are illustrated in Table \ref{tbl:settings}. The ACOPF
problem is formulated by using the model from the rosetta-opf project
\cite{rosetta-opf}, which is based on the models in PowerModels.jl \cite{8442948}, and the test cases are obtained from the pglib-opf
repository \cite{babaeinejadsarookolaee2019power}. We have selected
the goc and pegase cases because they contain large-scale instances.  The
external packages are called from Julia, through thin wrapper
packages, such as Ipopt.jl and AmplNLWriters.jl. A tolerance of
$10^{-4}$ is set for MadNLP.jl and Ipopt solvers, with other solver
options adjusted to ensure a fair comparison across different
solvers. The results can be reproduced with the script available at
\url{https://github.com/sshin23/opf-on-gpu}.

@@

\paragraph*{Convergence Pattern}
When comparing the solvers' performance in terms of the
IPM iteration counts, MadNLP.jl is as
efficient as the state-of-the-art solver Ipopt. The IPM iteration
count is nearly the same as that of Ipopt for achieving the same level
of accuracy in the final solution (see Table \ref{tbl:quality}).
{\color{blue} Note that
the constraint violation is not strictly less than the tolerance in all configurations
as relative (scaled) constraint violation is used for the convergence criteria.}
This
suggests that running mathematically equivalent operations on GPUs (by
using a shared code base in high-level abstractions) can yield a similar
degree of effectiveness in terms of IPM convergence.

@@

\paragraph*{Performance of the Condensed-Space IPM}
We next discuss the linear solver time. While the speedup achieved by
linear solvers is only moderate, this has a high impact on the overall
speedup because the linear solver time constitutes a significant
portion of the total solution time. {\color{blue} Figure
\ref{fig:speedup} reveals that linear algebra is the only
computational bottleneck for large-scale instances. While derivative
evaluation and solver internal computation could achieve a 10x speedup,
the linear algebra part can only achieve a 5x speedup even for the
largest case. For case24464\_goc, the GPU linear solver performance
was close to that of the CPU solver.  The investigation of under which
circumstances cuSOLVER is more effective warrants further research.}

@@

\paragraph*{Alternative Linear Solvers}
While cuSOLVER has been effective for solving the condensed KKT
systems using LU factorization, Cholesky factorization holds promise
for better performance due to lower computational complexity and the
ability to reveal the inertia of the KKT system. We are
interested in exploring other linear solver options, such as CHOLMOD
\cite{chen2008algorithm}, Baspacho \cite{pineda2022theseus}, and HyKKT
\cite{regev2023hykkt}.
